International Relations are a Social Network:\\ A Latent Network Approach for Global Politics

### 2017-10-17 paper/2_model.tex
more hand holding in model

The AME model combines the social relations regression model (SRRM) to account for nodal and dyadic dependencies and the latent factor model for third order dependencies.\footnote{For details on the SRRM see: for details on this model see \citet{li:loken:2002,dorff:minhas:2016}. An earlier version of the latent factor model (LFM) used in AME is presented as the general bilinear mixed effects (GBME) model in \citet{hoff:ward:2004}. The GBME model is more limited in the types of dependence patterns that it can capture due to differences in the type of matrix decomposition approach utilized.} The AME model is specified as follows:

\begin{align}
	\begin{aligned}
		y_{ij} = f(\theta_{ij}) &\text{, where } \\
		\theta_{ij} =& \bm\beta_{d}^{\top} \mathbf{X}_{ij} + \bm\beta_{s}^{\top} \mathbf{X}_{i} + \bm\beta_{r}^{\top} \mathbf{X}_{j} + a_{i} + b_{j}  + \textbf{u}_{i}^{\top} \textbf{D} \textbf{v}_{j} + \epsilon_{ij} \\ 
	\label{eqn:ame}
	\end{aligned}
\end{align}

where $y_{ij,t}$ captures the interaction between actor $i$ (the sender) and $j$ (the receiver). We use a Bayesian probit framework, in which we model a latent variable, $\theta_{ij}$, using first a set of exogenous dyadic ($\bm\beta_{d}^{\top} \mathbf{X}_{ij}$), sender ($\bm\beta_{s}^{\top} \mathbf{X}_{i}$), and receiver covariates ($\bm\beta_{r}^{\top} \mathbf{X}_{j}$); and then parameters from the SRRM and LFM. $a_{i}$ and $b_{j}$ in Equation~\ref{eqn:ame} represent sender and receiver random effects incorporated from the SRRM framework:

@@

The LFM contribution to the AME comes in the multiplicative term: $\textbf{u}_{i}^{\top} \textbf{D} \textbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third order interdependencies \citep{minhas:etal:2016:arxiv}. The representation of third order interdepedencies is accomplished by a process similar to computing the singular value decomposition (SVD) of the observed network. When taking the SVD we factorize our observed network into the product of three matrices: $\textbf{U}, \textbf{D}, \text{ and }, \textbf{V}$. This provides us with a low-dimensional representation of our original network in terms of third order dependence patterns arising in the network.\footnote{The dimensions of $\textbf{U}$ and $\textbf{V}$ are $n \times K$ and $\textbf{D}$ is a $K \times K$ diagonal matrix.} Specifically, values in $\textbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network, or more simply put how similar actors are in terms of who they are initiating, for example, conflict with. $\hat{\textbf{u}}_{i} \approx \hat{\textbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\textbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\textbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both the presence and absence of homophily.}

>>>>

The AME model combines the social relations regression model (SRRM) to account for nodal and dyadic dependencies and the latent factor model (LFM) for third order dependencies.\footnote{For details on the SRRM see: for details on this model see \citet{li:loken:2002,dorff:minhas:2016}. An earlier version of the LFM used in AME is presented as the general bilinear mixed effects (GBME) model in \citet{hoff:ward:2004}. The GBME model is more limited in the types of dependence patterns that it can capture due to the formulation of the matrix decomposition procedure.} The AME model is specified as follows:

\begin{align}
	\begin{aligned}
		y_{ij} \;=\; f(\theta_{ij}) &\text{, where } \\
		\theta_{ij} \;=\;& \bm\beta_{d}^{\top} \mathbf{X}_{ij} + \bm\beta_{s}^{\top} \mathbf{X}_{i} + \bm\beta_{r}^{\top} \mathbf{X}_{j} + \text{\qquad(Exogenous parameters)} \\
		& a_{i} + b_{j} + \epsilon_{ij} + \text{\qquad\qquad\qquad\quad(SRRM parameters)} \\
		& \textbf{u}_{i}^{\top} \textbf{D} \textbf{v}_{j}  \text{\qquad\qquad\qquad\qquad\qquad\;(LFM parameters)} \\ 
	\label{eqn:ame}
	\end{aligned}
\end{align}

where $y_{ij,t}$ captures the interaction between actor $i$ (the sender) and $j$ (the receiver). We use a Bayesian probit framework, in which we model a latent variable, $\theta_{ij}$, using first a set of exogenous dyadic ($\bm\beta_{d}^{\top} \mathbf{X}_{ij}$), sender ($\bm\beta_{s}^{\top} \mathbf{X}_{i}$), and receiver covariates ($\bm\beta_{r}^{\top} \mathbf{X}_{j}$);

this is where most people end.

and then parameters from the SRRM and LFM. $a_{i}$ and $b_{j}$ in Equation~\ref{eqn:ame} represent sender and receiver random effects incorporated from the SRRM framework:

@@

The LFM contribution to the AME comes in the multiplicative term: $\textbf{u}_{i}^{\top} \textbf{D} \textbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third order interdependencies \citep{minhas:etal:2016:arxiv}. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When taking the SVD we factorize our observed network into the product of three matrices: $\textbf{U}, \textbf{D}, \text{ and }, \textbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\textbf{U}$ and $\textbf{V}$ are $n \times K$ and $\textbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\textbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network, or more simply put how similar actors are in terms of who they are initiating, for example, conflict with. $\hat{\textbf{u}}_{i} \approx \hat{\textbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\textbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\textbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both the presence and absence of homophily.}

By integrating the SRRM and LFM into a Bayesian probit framework, we are able to account for the underlying structure in dyadic data that if left unestimated would complicate any inferences we draw
